---
title: "Markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Script information

<Description goes here>

This script was developed using R version 3.4.3 (64-bits).

#Setup
```{r}
#Set working directory.
setwd("/mnt/virdir/Backup/tjonkman/epigenetic-clocks/")

#Set library path.
.libPaths("/mnt/virdir/Backup/tjonkman/Rlibs")

#Load all necessary libraries.
library(BBMRIomics)
library(irlba)
library(ggfortify)
library(reshape2)
library(ggplot2)
library(cate)
library(ggbiplot)
library(foreach)
library(doParallel)

#Register 10 cores as parallel workers.
registerDoParallel(cores=10)

options(stringsAsFactors = T)
```

#Load data.
```{r}
load("Output/Output_03a_twas_data.RData")
```

#
#
#Clock CpGs
#
#


#Prepare data.
```{r}
mvalues <- mvalues.clocks

#Remove problem samples.
#There are 2 covariates which contain certain levels causing singularities if they are included into the model:
#flowcell_number: levels "30a" and "30b".
#sample_plate: levels "OV0192DNA001" and "OV0192DNA002".

idx <- which(mvalues$flowcell_num == "30a" | mvalues$flowcell_num == "30b" | mvalues$sample_plate == "OV0192DNA001" | mvalues$sample_plate == "OV0192DNA002")

#Now remove them.
mvalues <- mvalues[,-idx]

#Also remove the level from the factor variable(s) by re-declaring the factor (removes all unused levels).
mvalues$flowcell_num <- factor(mvalues$flowcell_num)
mvalues$sample_plate <- factor(mvalues$sample_plate)

mvalues
```

#Prepare CATE.
```{r}
#As an initialization step, define an empty "cpg" object (this will be filled in later for each iteration of the TWAS).
mvalues$cpg <- numeric(length = ncol(mvalues))

#Add the formula to the SummarizedExperiment.
metadata(mvalues)$formula <- ~ cpg + biobank_id + sampling_age + PredictedSex + baso_perc_imputed + eos_perc_imputed + lymph_perc_imputed + mono_perc_imputed + sample_plate + sentrix_position + flowcell_num

#Select covariates without NAs.
covariates <- get_all_vars(metadata(mvalues)$formula, data=colData(mvalues))
nas <- apply(covariates, 1, anyNA)
mvalues <- mvalues[, !nas]
mvalues

#Make a model matrix.
design <- model.matrix(metadata(mvalues)$formula, data=colData(mvalues))

#Create a matrix of the assays.
M.cate <- t(assay(mvalues))
#M.cate[1:10,1:10]

#Match RNA-seq samples to the DNA-methylation samples.
mvalues
idx <- match(colnames(mvalues), colnames(RIN.counts))
RIN.counts <- RIN.counts[,idx]
dim(RIN.counts)

#Transpose the RNA-seq data before running the model, so that it is in the right format (rownames should be the sample names).
count.cate <- t(RIN.counts)

dim(count.cate)

#Define TWAS using CATE to model gene expression, using the methylation of CpGs as covariates and estimating both known and estimated unknown confounders.
TWAS <- function(i){
  
  #Check progress
  print(paste0("CpG ", i, ": ", rownames(mvalues)[i]))
  
  #Select 1 CpG, and add it to the design matrix as the primary variable.
  cpg <- M.cate[,i]
  design[,2] <- cpg
  
  #Run the TWAS for this CpG.
  twas <- tryCatch({
    results.cate <- cate.fit(design[,2,drop=FALSE], design[,-2], count.cate, r = 5, calibrate=FALSE, fa.method="ml", adj.method = "naive")
    }, error = identity)
  if(is.null(twas$message)){
    
    #Extract the beta-estimates and p-values, and save them as a dataframe.
    dat <- as.data.frame(cbind(results.cate$beta, results.cate$beta.t, results.cate$beta.p.value))
    colnames(dat) <- c("beta", "t-statistic", "p-value")
    
  } else {
    #make output structure a dataframe.
    dat <- data.frame("CpG_name" = rownames(mvalues)[1], "Error" = "Didn't converge")
  }
  return(dat)
}
```

#Run CATE.
```{r}
idx <- 1:ncol(M.cate)
CATE.clocks <- foreach(i = idx, .errorhandling = 'stop') %dopar% TWAS(i)
```

#Check output.
```{r}
idx <- 1:length(CATE.clocks)

test <- foreach(i = idx, .combine = rbind) %do% dim(CATE.clocks[[i]])
table(test[,1] == ncol(count.cate)) #Should be 100% TRUE. If this is FALSE for some CpGs, the TWAS went wrong for those.
```

#Save output.
```{r}
save(CATE.clocks, file = "Output/Output_07a_CATE_clocks_full.RData")

#Split the output into beta-values and p-values.
idx <- 1:length(CATE.clocks)

#Effect sizes.
es.clocks <- foreach(i = idx, .combine = cbind) %dopar% CATE.clocks[[i]][,1]
dim(es.clocks)
rownames(es.clocks) <- colnames(count.cate)
colnames(es.clocks) <- colnames(M.cate)

#t-statistics.
t.clocks <- foreach(i = idx, .combine = cbind) %dopar% CATE.clocks[[i]][,2]
dim(t.clocks)
rownames(t.clocks) <- colnames(count.cate)
colnames(t.clocks) <- colnames(M.cate)

#p-values (unadjusted).
p.clocks <- foreach(i = idx, .combine = cbind) %dopar% CATE.clocks[[i]][,3]
dim(p.clocks)
rownames(p.clocks) <- colnames(count.cate)
colnames(p.clocks) <- colnames(M.cate)

save(es.clocks, file = "Output/Output_07b_CATE_clocks_effect_sizes.RData")
save(t.clocks, file = "Output/Output_07c_CATE_clocks_t_statistics.RData")
save(p.clocks, file = "Output/Output_07d_CATE_clocks_p_values.RData")
```

#
#
#Reference CpGs
#
#

#Prepare data.
```{r}
rm(CATE.clocks)
mvalues <- mvalues.refset

#Remove problem samples.
#There are 2 covariates which contain certain levels causing singularities if they are included into the model:
#flowcell_number: levels "30a" and "30b".
#sample_plate: levels "OV0192DNA001" and "OV0192DNA002".

idx <- which(mvalues$flowcell_num == "30a" | mvalues$flowcell_num == "30b" | mvalues$sample_plate == "OV0192DNA001" | mvalues$sample_plate == "OV0192DNA002")

#Now remove them.
mvalues <- mvalues[,-idx]

#Also remove the level from the factor variable(s) by re-declaring the factor (removes all unused levels).
mvalues$flowcell_num <- factor(mvalues$flowcell_num)
mvalues$sample_plate <- factor(mvalues$sample_plate)

mvalues
```

#Prepare CATE.
```{r}
#As an initialization step, define an empty "cpg" object (this will be filled in later for each iteration of the TWAS).
mvalues$cpg <- numeric(length = ncol(mvalues))

#Add the formula to the SummarizedExperiment.
metadata(mvalues)$formula <- ~ cpg + biobank_id + sampling_age + PredictedSex + baso_perc_imputed + eos_perc_imputed + lymph_perc_imputed + mono_perc_imputed + sample_plate + sentrix_position + flowcell_num

#Select covariates without NAs.
covariates <- get_all_vars(metadata(mvalues)$formula, data=colData(mvalues))
nas <- apply(covariates, 1, anyNA)
mvalues <- mvalues[, !nas]
mvalues

#Make a model matrix.
design <- model.matrix(metadata(mvalues)$formula, data=colData(mvalues))

#Create a matrix of the assays.
M.cate <- t(assay(mvalues))
#M.cate[1:10,1:10]

#Match RNA-seq samples to the DNA-methylation samples.
mvalues
idx <- match(colnames(mvalues), colnames(RIN.counts))
RIN.counts <- RIN.counts[,idx]
dim(RIN.counts)

#Transpose the RNA-seq data before running the model, so that it is in the right format (rownames should be the sample names).
count.cate <- t(RIN.counts)

dim(count.cate)
```

#Define TWAS using CATE to model gene expression, using the methylation of CpGs as covariates and estimating both known and estimated unknown confounders.
(Omitted, is already done above)

#Run CATE.
```{r}
idx <- 1:ncol(M.cate)
CATE.refset <- foreach(i = idx, .errorhandling = 'stop') %dopar% TWAS(i)
```

#Check output.
```{r}
idx <- 1:length(CATE.refset)

test <- foreach(i = idx, .combine = rbind) %do% dim(CATE.refset[[i]])
table(test[,1] == ncol(count.cate)) #Should be 100% TRUE. If this is FALSE for some CpGs, the TWAS went wrong for those.
```

#Save output.
```{r}
save(CATE.refset, file = "Output/Output_07e_CATE_refset_full.RData")

#Split the output into beta-values and p-values.
idx <- 1:length(CATE.refset)

#Effect sizes.
es.refset <- foreach(i = idx, .combine = cbind) %dopar% CATE.refset[[i]][,1]
dim(es.refset)
rownames(es.refset) <- colnames(count.cate)
colnames(es.refset) <- colnames(M.cate)

#t-statistics.
t.refset <- foreach(i = idx, .combine = cbind) %dopar% CATE.refset[[i]][,2]
dim(t.refset)
rownames(t.refset) <- colnames(count.cate)
colnames(t.refset) <- colnames(M.cate)

#p-values (unadjusted).
p.refset <- foreach(i = idx, .combine = cbind) %dopar% CATE.refset[[i]][,3]
dim(p.refset)
rownames(p.refset) <- colnames(count.cate)
colnames(p.refset) <- colnames(M.cate)

save(es.refset, file = "Output/Output_07f_CATE_refset_effect_sizes.RData")
save(t.refset, file = "Output/Output_07g_CATE_refset_t_statistics.RData")
save(p.refset, file = "Output/Output_07h_CATE_refset_p_values.RData")
```

#Session info.
```{r}
sessionInfo()
```