---
title: "Markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Script information

This script was written to prepare Genomic Ranges objects of 4 epigenetic clocks:
- Hannum Bld: 71 CpGs (Hannum G et al. Genome-wide methylation profiles reveal quantitative views of human aging rates. Mol Cell. 2013)
- Horvath MT: 353 CpGs (aDMCs) (Horvath S. DNA methylation age of human tissues and cell types. Genome Biol. 2013)
- Horvath Skn/Bld: 391 CpGs (Horvath S et al. Epigenetic Clock for Skin and Blood Cells Applied to Hutchinson Gilford Progeria Syndrome and ex vivo Studies. Aging. 2018)
- Zhang Bld/Slv: 514 aDMCs included in Zhang et al's epigenetic clock (Zhang Q et al. Improved precision of epigenetic clock estimates across tissues and its implication for biological ageing. Genome Med. 2019)

#Setup
```{r}
#Set working directory.
setwd("/home/tjonkman/researchdrive/tjonkman/epigenetic-clocks")

#set library path.
.libPaths("/home/tjonkman/researchdrive/tjonkman/epigenetic-clocks/Packages")

#Load all necessary libraries.
library(GenomicRanges)
library(ggplot2)

options(stringsAsFactors = F)
```

#Load data.
```{r}
#Load 450K array annotation.
load("Output/Output_01b_450K_annotation.RData")

#Load 4 aDMP-sets.
hannum <- read.csv("Input/Input_02b_hannum_cpgs.csv")
horvath <- read.csv("Input/Input_02a_horvath_cpgs.csv")
skinblood <- read.csv("Input/Input_02d_skinblood.cpgs.csv")
zhang <- read.csv("Input/Input_02c_zhang_cpgs.csv")
```

#Prepare GRanges objects of all 4 CpG sets, and also make a GRanges object which contains the CpGs of all 4 clocks.
```{r}
hannum <- sort(sortSeqlevels(ann450K[hannum$CpG]))
horvath <- sort(sortSeqlevels(ann450K[horvath$CpG]))
skinblood <- sort(sortSeqlevels(ann450K[na.omit(match(skinblood$CpG, names(ann450K)))]))
zhang <- sort(sortSeqlevels(ann450K[zhang$CpG]))

#Union of the 4 clocks.
all.clocks <- unique(c(hannum, horvath, skinblood, zhang))
all.clocks$hannum <- F
all.clocks$horvath <- F
all.clocks$skinblood <- F
all.clocks$zhang <- F

#Add the information on which clocks contain each of the CpGs (NB: some CpGs are present in multiple clocks).
all.clocks$hannum[match(names(hannum), names(all.clocks))] <- T
all.clocks$horvath[match(names(horvath), names(all.clocks))] <- T
all.clocks$skinblood[match(names(skinblood), names(all.clocks))] <- T
all.clocks$zhang[match(names(zhang), names(all.clocks))] <- T
all.clocks

#Make a reference set of non-clock CpGs. This will be used as a background for enrichment testing.
idx <- match(names(all.clocks), names(ann450K))
non.clock <- ann450K[-idx]
```

#Save the CpG sets.
```{r}
save(hannum, horvath, skinblood, zhang, all.clocks, file = "Output/Output_02a_clocks.RData")
save(non.clock, file = "Output/Output_02b_non.clock.RData")
```

#Session info.
```{r}
sessionInfo()
```