---
title: "Markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Script information

Link to the bacon package article: https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1131-9

This script was developed using R version 3.4.3 (64-bits).

#Setup
```{r}
#Set working directory.
setwd("/mnt/virdir/Backup/tjonkman/epigenetic-clocks/")

#Set library path.
.libPaths("/mnt/virdir/Backup/tjonkman/Rlibs")

#Load all necessary libraries.
library(bacon)

#Register 10 cores as parallel workers.
register(MulticoreParam(10, log=TRUE))

options(stringsAsFactors = F)
```

#Run bacon on t-statistics obtained from the TWAS, in order to correct for any residual inflation and bias of the t-statistics.
#NB: bacon will be run separately for each CpG. This is because CATE was run using a separate model for each CpG, using that CpG to predict the expression of all 14,000 genes. In other words, each CpG represents a single related set of t-statistics, generated from the same linear model.
```{r}
#Load matrix of t-statistics of clocks and reference CpGs.
load("Output/Output_07b_CATE_clocks_effect_sizes.RData")
load("Output/Output_07c_CATE_clocks_t_statistics.RData")

#Set random seed (for reproducibility).
set.seed(1)

#Run BACON.
bc.clocks <- bacon(effectsizes = es.clocks, teststatistics = t.clocks)

#Extract the BACON-adjusted t-statistics and p-values for both clocks and refset.
#es.clocks <- es(bc.clocks)
#se.clocks <- se(bc.clocks)
t.clocks <- tstat(bc.clocks)
p.clocks <- pval(bc.clocks)
```

#Inspect output for clocks.
```{r}
#Inspect inflations and biases.
inflations <- inflation(bc.clocks)
summary(inflations)
hist(inflations, 100)

biases <- bias(bc.clocks)
summary(biases)
hist(biases, 100)

#
#Plot the distributions of the most and least inflated CpGs, and of the most and least biased CpGs.
#

#Least inflated (closest to 1)
test <- abs(1-inflations)
idx <- order(test, decreasing = F)
inflations[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Least inflated")
posteriors(bc.clocks, index = idx[1])

#Most inflation
idx <- order(inflations, decreasing = T)
inflations[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Most inflation")
posteriors(bc.clocks, index = idx[1])

#Most deflation
idx <- order(inflations, decreasing = F)
inflations[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Most deflation")
posteriors(bc.clocks, index = idx[1])


#Least biased
test <- abs(0-biases)
idx <- order(test, decreasing = F)
biases[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Least bias")
posteriors(bc.clocks, index = idx[1])

#Most negative bias
idx <- order(biases, decreasing = F)
biases[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Most negative bias")
posteriors(bc.clocks, index = idx[1])

#Most positive bias
idx <- order(biases, decreasing = T)
biases[idx[1:10]]
fit(bc.clocks, n=50, index = idx[1], xlim = c(-6,6), main = "Most positive bias")
posteriors(bc.clocks, index = idx[1])
```

#Save BACON output.
```{r}

#clocks
save(bc.clocks, file = "Output/Output_08a.0_bacon_full_clocks.RData")
save(t.clocks, file = "Output/Output_08a.1_bacon_t_clocks.RData")
save(p.clocks, file = "Output/Output_08a.2_bacon_p_clocks.RData")
```

#Session info.
```{r}
sessionInfo()
```