---
title: "Markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Script information

This script was written to test if aDMPs (hannum) are enriched for certain annotations. This is done by caluculating the odds ratios of them falling into certain annotations compared to a random set of reference CpGs (ref.hannum).

This script was developed using R version 3.4.3 (64-bits).

#Setup
```{r}
#Set working directory.
setwd("/mnt/virdir/Backup/tjonkman/epigenetic-clocks/")

#Set library path.
.libPaths("/mnt/virdir/Backup/tjonkman/Rlibs")

#Load all necessary libraries.
library(GenomicRanges)
library(gplots)
```

#Prepare data.
```{r}
#Load 450K annotation, hannum CpGs, and reference set.
load("Output/Output_01b_450K_annotation.RData")
load("Output/Output_03b_clocks_full.RData")
load("Output/Output_03c_refset_full.RData")

idx <- colnames(values(all.clocks))[grep("H3K", colnames(values(all.clocks)))]
for(i in idx){
  values(all.clocks)[,i] <- factor(values(all.clocks)[,i])
  values(refset)[,i] <- factor(values(refset)[,i])
  values(non.clock)[,i] <- factor(values(non.clock)[,i])
}

#Add a vector indicating whether each CpG has a positive correlation with age.
all.clocks$age.cor.above.0 <- all.clocks$Age.cor > 0

hannum <- all.clocks[all.clocks$hannum]
horvath <- all.clocks[all.clocks$horvath]
skinblood <- all.clocks[all.clocks$skinblood]
zhang <- all.clocks[all.clocks$zhang]
```

#Declare 3 functions: 
- OddsRatios calculates odds ratios for a single feature.
- CallOddsRatios calls OddsRatios for every feature.
- PlotHeatmap plots these odds ratios in a heatmap.
```{r}
OddsRatios <- function(feature, set.of.interest, background.set, set.of.interest.name, background.set.name){
  
  #Create a matrix for storing the odds ratios and p-values.
  colnms <- c(paste0(background.set.name, ", not in element", sep = ""), 
              paste0(background.set.name, ", in element",  sep = ""), 
              paste0(set.of.interest.name, ", not in element", sep = ""), 
              paste0(set.of.interest.name, ", in element", sep = ""), 
              paste0("Odds ", background.set.name, sep = ""), 
              paste0("Odds ", set.of.interest.name, sep = ""), 
              "Odds ratio", 
              "p-value")
  rwnms <- levels(values(set.of.interest[,feature])[,1])
  dim.nms <- list(rwnms,colnms)
  ORtab <<- matrix(nrow = length(rwnms), ncol = length(colnms), dimnames = dim.nms)
  
  #Calculate odds ratios and p-values.
  
  for(i in 1:length(levels(values(background.set[,feature])[,1]))){
    
    n00 <- length(background.set[values(background.set[,feature])[,1] != levels(values(background.set[,feature])[,1])[i]])
    n01 <- length(background.set[values(background.set[,feature])[,1] == levels(values(background.set[,feature])[,1])[i]])
    n10 <- length(set.of.interest[values(set.of.interest[,feature])[,1] != levels(values(set.of.interest[,feature])[,1])[i]])
    n11 <- length(set.of.interest[values(set.of.interest[,feature])[,1] == levels(values(set.of.interest[,feature])[,1])[i]])
    
    Odds.background.set <- n01/n00 #Odds of background set being in a feature compared to not being in one.
    Odds.set.of.interest <- n11/n10 #Odds of set of interest being in a feature compared to not being in one.
    OR <- Odds.set.of.interest/Odds.background.set #Odds ratio of set of interest being in a feature, compared to background set.
    
    #Perform a Fisher's exact test to test for significance of the ORs. NB: some p-values will be so low that they will be returned as 0.
    dat <- matrix(c(n00, n01, n10, n11), byrow=T, ncol=2)
    colnames(dat) <- c("Not in element", "In element")
    rownames(dat) <- c("reference CpG", "aVMC")
    fit <- fisher.test(dat)
    
    #Save the values into ORtab.
    ORtab[i,] <<- c(n00, n01, n10, n11, Odds.background.set, Odds.set.of.interest, OR, fit$p.value)
  }
  
  # #Set odds ratio to 0 if the p-values are not significant (p-Bonferroni < 0.05).
  # for(i in 1:nrow(ORtab)){
  #   if(ORtab[i,8] > (0.05)) {ORtab[i,7] <<- 1}
  # }
  return(ORtab)
}

CallOddsRatios <- function(set.of.interest, background.set, set.of.interest.name, background.set.name) {
  
  #CGI features.
  OddsRatios(feature = "CGI_Feature", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.CGI <<- ORtab
  
  #Chromatin state.
  OddsRatios(feature = "Chromatin.state.abbr", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Chromatin <<- ORtab
  
  #Histone modifications.
  OR.Hist.merge <<- data.frame(matrix(NA, nrow = 6, ncol = 8))
  rownames(OR.Hist.merge) <<- c("H3K4me1", "H3K4me3", "H3K27ac", "H3K36me3", "H3K9me3", "H3K27me3")
  colnames(OR.Hist.merge) <<- colnames(ORtab)
  
  #H3K4me1.
  OddsRatios(feature = "H3K4me1", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[1,] <<- ORtab[2,]
  
  #H3K4me3.
  OddsRatios(feature = "H3K4me3", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[2,] <<- ORtab[2,]
  
  #H3K27ac.
  OddsRatios(feature = "H3K27ac", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[3,] <<- ORtab[2,]
  
  #H3K36me3.
  OddsRatios(feature = "H3K36me3", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[4,] <<- ORtab[2,]
  
  #H3K9me3.
  OddsRatios(feature = "H3K9me3", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[5,] <<- ORtab[2,]
  
  #H3K27me3.
  OddsRatios(feature = "H3K27me3", set.of.interest, background.set, set.of.interest.name, background.set.name)
  OR.Hist.merge[6,] <<- ORtab[2,]
}

PlotHeatmap <- function(file.path){
  #hannum
CallOddsRatios(set.of.interest = hannum, background.set = non.clock, set.of.interest.name = "hannum", background.set.name = "non.clock")
OR.hannum <- as.data.frame(rbind(OR.CGI, OR.Hist.merge, OR.Chromatin))
# OR.hannum$p.adj <- p.adjust(OR.hannum$`p-value`, method = "bonferroni")
OR.hannum$p.adj <- OR.hannum$`p-value` * 9
OR.hannum$p.adj[OR.hannum$p.adj > 1] <- 1
OR.hannum$`Odds ratio`[OR.hannum$p.adj >= 0.05] <- 1
# OR.hannum

#horvath
CallOddsRatios(set.of.interest = horvath, background.set = non.clock, set.of.interest.name = "horvath", background.set.name = "non.clock")
OR.horvath <- as.data.frame(rbind(OR.CGI, OR.Hist.merge, OR.Chromatin))
# OR.horvath$p.adj <- p.adjust(OR.horvath$`p-value`, method = "bonferroni")
OR.horvath$p.adj <- OR.horvath$`p-value` * 9
OR.horvath$p.adj[OR.horvath$p.adj > 1] <- 1
OR.horvath$`Odds ratio`[OR.horvath$p.adj >= 0.05] <- 1
# OR.horvath

#skinblood
CallOddsRatios(set.of.interest = skinblood, background.set = non.clock, set.of.interest.name = "skinblood", background.set.name = "non.clock")
OR.skinblood <- as.data.frame(rbind(OR.CGI, OR.Hist.merge, OR.Chromatin))
# OR.skinblood$p.adj <- p.adjust(OR.skinblood$`p-value`, method = "bonferroni")
OR.skinblood$p.adj <- OR.skinblood$`p-value` * 9
OR.skinblood$p.adj[OR.skinblood$p.adj > 1] <- 1
OR.skinblood$`Odds ratio`[OR.skinblood$p.adj >= 0.05] <- 1
# OR.skinblood

#zhang
CallOddsRatios(set.of.interest = zhang, background.set = non.clock, set.of.interest.name = "zhang", background.set.name = "non.clock")
OR.zhang <- as.data.frame(rbind(OR.CGI, OR.Hist.merge, OR.Chromatin))
# OR.zhang$p.adj <- p.adjust(OR.zhang$`p-value`, method = "bonferroni")
OR.zhang$p.adj <- OR.zhang$`p-value` * 9
OR.zhang$p.adj[OR.zhang$p.adj > 1] <- 1
OR.zhang$`Odds ratio`[OR.zhang$p.adj >= 0.05] <- 1
# OR.zhang

#Merge the odds ratio tables into one table, and transform the results onto a log2-scale for visualization.
OR.merge <- cbind(OR.hannum$`Odds ratio`, OR.horvath$`Odds ratio`, OR.skinblood$`Odds ratio`, OR.zhang$`Odds ratio`)
rownames(OR.merge) <- rownames(OR.hannum)
clock.names <- c("Hannum Bld", "Horvath PT", "Horvath Skn/Bld", "Zhang Bld/Slv")
colnames(OR.merge) <- clock.names
OR.merge <- log2(OR.merge)
OR.merge

#Add two empty rows (called spacer1 and spacer2) to the data for layout purposes.
spacer1 <- data.frame(NA,NA,NA,NA , row.names = "")
colnames(spacer1) <- clock.names
spacer2 <-  data.frame(NA,NA,NA,NA , row.names = " ")
colnames(spacer2) <- clock.names
plot.data <- rbind(OR.merge[1:3,], spacer1, OR.merge[4:9,], spacer2, OR.merge[10:24,])
plot.data <- as.matrix(plot.data)
plot.data

#Remove the number codes in the chromatin states.
split.names <- strsplit(rownames(plot.data)[12:26], "_")
split.names <- as.character(lapply(split.names, function(x){x[2]}))
rownames(plot.data)[12:26] <- split.names
plot.data

my.palette <- colorRampPalette(c("#0000cc", "#2222ff", "#eeeeee", "#ff2222", "#cc0000"))(n = 11)
my.breaks <- seq(-3, 3, length = 12)

#Plot the odds ratios in a heatmap.
png(file = file.path, width = 2, height = 6, unit = "in", res = 300)
heatmap.2(plot.data, Colv = F, Rowv = F, dendrogram = "none", trace = "none", col = my.palette, breaks = my.breaks, density.info = "none", key = F, keysize = 0.1, margins = c(7.5, 5.5), cexRow = 1, cexCol = 1, na.color=par("bg"))
dev.off()
}
```

#Calculate genomic enrichments for all clocks, and plot them in a heatmap.
```{r}
PlotHeatmap(file.path = "Output/Plot_06a_genomic_enrichment_heatmap_all.png")
```

#Repeat the odds ratio calculation, but now for CpGs which correlate either positively or negatively with age.
```{r}
#Positively-correlated CpGs.
hannum <- all.clocks[(all.clocks$hannum & all.clocks$age.cor.above.0)]
horvath <- all.clocks[(all.clocks$horvath & all.clocks$age.cor.above.0)]
skinblood <- all.clocks[(all.clocks$skinblood & all.clocks$age.cor.above.0)]
zhang <- all.clocks[(all.clocks$zhang & all.clocks$age.cor.above.0)]

PlotHeatmap(file.path = "Output/Plot_06b_genomic_enrichment_heatmap_positive.png")

#Negatively-correlated CpGs.
hannum <- all.clocks[(all.clocks$hannum & !all.clocks$age.cor.above.0)]
horvath <- all.clocks[(all.clocks$horvath & !all.clocks$age.cor.above.0)]
skinblood <- all.clocks[(all.clocks$skinblood & !all.clocks$age.cor.above.0)]
zhang <- all.clocks[(all.clocks$zhang & !all.clocks$age.cor.above.0)]

PlotHeatmap(file.path = "Output/Plot_06c_genomic_enrichment_heatmap_negative.png")
```

#Session info.
```{r}
sessionInfo()
```